---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

Inspired by [Emma Baccellieri's](https://twitter.com/emmabaccellieri) article ["White Sox' Yasmani Grandal Is Doing the Most With the Least but How Is He Making It Work?"](https://www.si.com/mlb/2021/06/02/yasmani-grandal-the-opener). Here's my attempt to visualize Yasmani Grandal's absurd 2021 stat line.

This readme contains all of the analysis and functions to automate the data prep and wOBA component calculation steps can be found in 99_woba-functions.R

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load the stats

We're going to grab Yasmani Grandal's stats from baseball-reference. You could do this with an html handler, but it's also pretty easy to just go to [his br page](https://www.baseball-reference.com/players/g/grandya01.shtml) and grab his stat line in csv format. Conveniently, he's never played for multiple teams in a single season.

```{r yasmani_stats, message=FALSE}
library(tidyverse)

# a.o 6/1
# https://www.baseball-reference.com/players/g/grandya01.shtml
yasmani_stats <- read_csv(
  "Year,Age,Tm,Lg,G,PA,AB,R,H,2B,3B,HR,RBI,SB,CS,BB,SO,BA,OBP,SLG,OPS,OPS+,TB,GDP,HBP,SH,SF,IBB,Pos,Awards
2012,23,SDP,NL,60,226,192,28,57,7,1,8,36,0,0,31,39,.297,.394,.469,.863,143,90,8,1,0,2,1,2/H,
2013,24,SDP,NL,28,108,88,13,19,8,0,1,9,0,0,18,18,.216,.352,.341,.693,102,30,1,1,0,1,2,2/H3,
2014,25,SDP,NL,128,443,377,47,85,19,1,15,49,3,0,58,115,.225,.327,.401,.728,111,151,7,2,0,6,1,23H/D,
2015,26,LAD,NL,115,426,355,43,83,12,0,16,47,0,1,65,92,.234,.353,.403,.756,112,143,16,2,1,3,1,2H/3,AS
2016,27,LAD,NL,126,457,390,49,89,14,1,27,72,1,3,64,116,.228,.339,.477,.816,118,186,11,2,0,1,1,*2H/3,MVP-22
2017,28,LAD,NL,129,482,438,50,108,27,0,22,58,0,1,40,130,.247,.308,.459,.767,101,201,10,0,1,3,0,*2H,
2018,29,LAD,NL,140,518,440,65,106,23,2,24,68,2,1,72,124,.241,.349,.466,.815,121,205,12,3,0,3,1,*2H/3,
2019,30,MIL,NL,153,632,513,79,126,26,2,28,77,5,1,109,139,.246,.380,.468,.848,119,240,16,5,0,5,2,*23/HD,ASMVP-15
2020,31,CHW,AL,46,194,161,27,37,7,0,8,27,0,0,30,58,.230,.351,.422,.773,112,68,4,1,0,2,0,2/D3H,
2021,32,CHW,AL,37,144,99,22,13,2,0,6,18,0,0,42,40,.131,.385,.333,.718,107,33,6,0,1,2,0,*2/3H,")

```

# Calculate wOBA

I've decided to do this analysis using [Fangraphs's wOBA](https://library.fangraphs.com/offense/woba/) in order to see how different plate outcomes contribute to the total. In simple terms, the formula includes positive outcomes in the numerator and negative outcomes in the denominator.
```{r wOBA weights}
wOBA_numer_weights <- tribble(
  ~stat, ~numer_weight, ~denom_weight,
  # the method only counts unintentional walks, so they'll need to be extracted first
  "uBB",  0.69,          1,
  "HBP",  0.72,          1,
  "SF",   0,             1,
  "AB",   0,             1,
  "1B",   0.89,          0,
  "2B",   1.27,          0,
  "3B",   1.62,          0,
  "HR",   2.10,          0
)

```

A small amount of prep is needed to get the source data into the right format for wOBA computation. Specifically, we'll need to first compute unintentional walks (uBB = BB - IBB) and singles (1B = H - 2B - 3B - HR), and then pivot the data into a longer format, which will make computation easier.

```{r data prep}
identifier_cols <- c("Year", "Age", "Tm", "Lg", "G")
weight_cols <- wOBA_numer_weights$stat

yasmani_stats_prepped <- yasmani_stats %>% 
  mutate(uBB = BB - IBB,
         `1B` = H - `2B` - `3B` - HR) %>% 
  # we only want to keep columns that matter for wOBA calculation plus a few identifiers
  select(all_of(identifier_cols), all_of(weight_cols)) %>% 
  pivot_longer(all_of(weight_cols), names_to = "stat", values_to = "total")

```

Once we've got the data in the right format, we can compute the numerator and denominator components of each stat, each stat's contribution to the overall wOBA, and overall value of wOBA itself.
```{r calculate woba}
yasmani_components <- yasmani_stats_prepped %>% 
  # attach the weights to the stat dataset
  left_join(wOBA_numer_weights, by = "stat") %>% 
  # apply the weights to calculate the numerator and denominator component of each outcome
  mutate(numer_component = total * numer_weight,
         denom_component = total * denom_weight) %>% 
  # and compute the total denominator (~ plate appearances)
  with_groups(Year, mutate, 
              denom_year = sum(denom_component),
              wOBA_year  = sum(numer_component) / denom_year) %>% 
  mutate(wOBA_component = numer_component / denom_year) %>% 
  select(all_of(identifier_cols), stat, total, 
         wOBA_component, wOBA_year, numer_component, denom_component, denom_year)

```

# Plotting wOBA components

First, a quick plot of Yasmani Grandal's wOBA over the years against some thresholds from [FanGraph's glossary](https://library.fangraphs.com/offense/woba/):
```{r wOBA plot, echo=FALSE}
yasmani_wOBA_only <- yasmani_components %>% 
  with_groups(all_of(identifier_cols), summarize,
              numer = sum(numer_component),
              denom = sum(denom_component)) %>% 
  mutate(wOBA = numer / denom, .keep = "unused")


# a few other things
wOBA_thresholds <- tribble(
  ~Rating,         ~wOBA,
  "Excellent",     .400,
  "Great",         .370,
  "Above Average", .340,
  "Average",       .320,
  "Below Average", .310,
  "Poor",          .300,
  "Awful",         .290
)

years <- yasmani_wOBA_only$Year
comp_color <- "grey40"

# grab a tiny bit of info about his best year
best_year <- yasmani_wOBA_only %>% 
  filter(G > 60) %>% 
  slice_max(wOBA, n = 1) %>% 
  mutate(wOBA_format = scales::number(wOBA, accuracy = 0.001),
         by_lab = glue::glue("Career best: {wOBA_format}\nin {Year} with {Tm}"))

wOBA_plot_basics <- ggplot(mapping = aes(x = Year, wOBA)) +
  scale_x_continuous(breaks = years, minor_breaks = NULL) +
  theme_bw() +
  theme(panel.grid = element_blank())

wOBA_plot_basics +
  geom_hline(aes(yintercept = wOBA), 
             data = wOBA_thresholds, 
             color = comp_color) +
  geom_text(aes(label = Rating), 
            data = wOBA_thresholds, 
            x = max(years) + 1, 
            nudge_y = 0.003,
            color = comp_color) +
  geom_line(data = yasmani_wOBA_only) +
  geom_point(data = yasmani_wOBA_only) +
  geom_text(aes(label = by_lab), data = best_year, nudge_y = 0.01) +
  scale_y_continuous("wOBA", breaks = wOBA_thresholds$wOBA,
                     labels = scales::label_number(accuracy = 0.001)) +
  expand_limits(x = max(years + 1.5)) +
  ggtitle("Yasmani Grandal wOBA")
```

Then a first attempt to plot the impact that each hitting outcome had each year.
```{r components plot, echo=FALSE}

# only want the stats that positively contribute to wOBA, make it a factor
yasmani_components_forplot <- yasmani_components %>% 
  semi_join(filter(wOBA_numer_weights, numer_weight > 0), by = "stat") %>% 
  mutate(stat_ord = fct_relevel(stat, "uBB", "HBP", "1B", "2B", "3B", "HR")) %>% 
  rename(wOBA = wOBA_component)

colors <- c("#7570b3", "#d95f02", "#ffffcc", "#a1dab4", "#41b6c4", "#225ea8")

wOBA_plot_basics +
  geom_bar(aes(fill = stat_ord),
           data = yasmani_components_forplot,
           stat = "identity", position = "stack") +
  scale_y_continuous("wOBA", breaks = seq(0, 0.6, by = 0.05),
                     labels = scales::label_number(accuracy = 0.001)) +
  scale_fill_manual("Component", values = colors) +
  ggtitle("Yasmani Grandal wOBA Components",
          subtitle = "In 2021, Grandal has walked enough to replace almost all value from non-HR hits")
```

